---
title: "Arkansas School Enrollment: An Overview"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Arkansas School Enrollment: An Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)
```

This vignette provides an overview of Arkansas school enrollment data access through the `arschooldata` package.

## Getting Started

```{r load-packages}
library(arschooldata)
library(dplyr)
```

## Available Data

The package provides access to Arkansas school enrollment data from the Arkansas Department of Education. Use `get_available_years()` to see the available date range:
```{r available-years}
get_available_years()
```

## Fetching Data

Use `fetch_enr()` to download enrollment data for a single year:

```{r fetch-example, eval=FALSE}
# Fetch 2024 enrollment data
enr_2024 <- fetch_enr(2024)

# View the structure
head(enr_2024)
```

Use `fetch_enr_multi()` to download multiple years at once:

```{r fetch-multi-example, eval=FALSE}
# Fetch 5 years of data
enr_multi <- fetch_enr_multi(2020:2024)

# View year counts
table(enr_multi$end_year)
```

## Data Structure

The returned data includes district-level information from the Arkansas Annual Statistical Reports. Key columns include:

- District identifiers and names
- Average Daily Attendance (ADA)
- Financial and budget information

## Data Sources

Arkansas school data comes from the Arkansas Division of Elementary and Secondary Education (DESE):

- **Annual Statistical Reports**: District-level fiscal and enrollment summary data
- **ADE Data Center**: Detailed enrollment demographics (https://adedata.arkansas.gov/statewide/)

## Caching

By default, downloaded data is cached locally to avoid repeated downloads. Use `cache_status()` to check cached data:

```{r cache-status}
cache_status()
```

Clear the cache with `clear_cache()` if needed.

## More Information

For detailed enrollment demographics by race/ethnicity, grade level, and school, visit the Arkansas Department of Education Data Center directly at https://adedata.arkansas.gov/statewide/

---

*For more information about Arkansas education data, visit the [Arkansas Division of Elementary and Secondary Education](https://dese.ade.arkansas.gov/).*
